Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-01 manuscript.tex
changed title
intro edits, more in progress

\title{Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression}

@@

\lettrine[lines=2]{\color{color2}M}{}orphologically distinct sex chromosomes have evolved multiple times independently in both plant and animal kingdoms \citep{westergaard1958,ohno1967,bull1983,charlesworth1991}. Despite numerous biological differences between the kingdoms, X and Y chromosomes in both lineages have undergone similar genetic changes, suggesting that general evolutionary mechanisms might drive sex chromosome divergence \citep{charlesworth1978,charlesworth1996CB,charlesworth2000degeneration}. Investigating the shared features of plant and animal sex chromosomes, including the loss of recombination \citep{bergero2009}, the degeneration of Y chromosomes \citep{hough2014,bergero2015}, and the evolution of dosage compensation \citep{muyle2012,papadopulos2015}, thus provides an informative context for investigating the evolutionary forces driving sex chromosome evolution.

One fundamental difference between sex chromosomes and autosomes is their copy number in a population; whereas autosomes in diploid species are present in two copies in each sex, the X chromosome is present in two copies in females and only one copy in males. As a consequence, the effective population size ($N_{e}$) of the X chromosome is predicted to be 3/4 that of autosomes, whereas the $N_{e}$ of the male-limited Y chromosome is expected to equal 1/4 that of autosomes (assuming an equal number of reproducing females and males). Such differences in $N_{e}$ are predicted to directly affect relative levels of neutral polymorphism maintained on these chromosomes, which is proportional to the product of $N_{e}$ and the neutral mutation rate, $\mu$ \citep{Kimura1984}. Such variation in neutral polymorphism within the genome can have important consequences on patterns of DNA sequence evolution, including the effectiveness of both positive and negative selection \citep{charlesworth1987}.

@@

@@

@@

@@

\subsection*{Background Selection and Selective Sweeps}

@@

@@

@@

>>>>

\title{Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression}

@@

\lettrine[lines=2]{\color{color2}M}{}orphologically distinct sex chromosomes have evolved multiple times independently in both plant and animal kingdoms \citep{westergaard1958,ohno1967,bull1983,charlesworth1991}. Despite clear biological differences between these kingdoms, X and Y chromosomes in both lineages have undergone similar genetic changes. For example, in both groups the loss of recombination between X and Y chromosomes is associated with genetic degeneration and loss of genes on the Y chromosome \citep{hough2014,bergero2015}. In some species this has also led to the evoltion dosage compensation \citep{charlesworth1996CB,muyle2012,papadopulos2015}, and changes to levels of X and Y genetic diversity \citep{ellegren2011,bachtrog2013NRG}. The independent evolution of these phenomena in a broad range of species suggest that general evolutionary mechanisms may be involved, but inferring the causes of molecular evolution and patterns of polymorphism in regions that lack recombination is a difficult challenge for both theoreticians and experimentalists \citep{charlesworth1978,charlesworth1996CB,charlesworth2000degeneration}.

One fundamental difference between the X and Y chromosomes is that there are 1/3 as many Y-linked gene copies as X-linked ones in a diploid population. As a consequence, the neutrally-expectied effective population size ($N_{e}$) of the Y chromosome is predicted to be 1/4 that of the autosomes, whereas the $N_{e}$ for the X chromosome should be 3/4 that of autosomes (assuming an equal number of reproducing females and males). Such differences in $N_{e}$ are predicted to directly affect relative levels of neutral polymorphism maintained on these chromosomes, which is proportional to the product of $N_{e}$ and the neutral mutation rate, $\mu$ \citep{Kimura1984}. Such variation in neutral polymorphism within the genome can have important consequences on patterns of DNA sequence evolution, including the effectiveness of both positive and negative selection \citep{charlesworth1987}.

@@

%In species with male heterogamety, the high variance in male reproductive success associated with sexual selection could further reduce the e¡ective population size of the Y chromosome and (to a smaller degree) of the autosomes, relative to that of the X chromosome (Caballero 1995). Ne-values can be estimated from data on variation in silent DNA sequences, if mutation rates for neutral or nearly neutral changes are accurately known (Kimura 1983). Evidence for sex di¡erences in the variance of reproductive success can therefore be obtained by measuring levels of silent variability; strong sexual selection on males should bring the X-linked value close to, or even higher than, the auto- somal value (Charlesworth 1996b).

@@

%distinct but related, process is the `weak selec- tion Hill^Robertson’ e¡ect (McVean & Charlesworth 2000). Closely linked selected alleles interfere with each other, inhibiting the spread of favourable alleles and the elimination of deleterious ones (Fisher 1930, chapter5; Muller 1932; Hill & Robertson 1966; Felsenstein 1974; Birky & Walsh 1988).

@@

% Because the mean size of a eukaryote gene is about 500 codons, and a Drosophila or plant chromosome might have about 2000^3000 genes, the number of third coding p ositions at which silent muta- tions can occur on a proto-Ychromosome is of the order of 106, ten times the number for the most extreme example shownin¢gure1.

@@

%Evidence for sex di¡erences in the variance of reproductive success can therefore be obtained by measuring levels of silent variability; strong sexual selection on males should bring the X-linked value close to, or even higher than, the auto- somal value (Charlesworth 1996b).
\subsection*{Background Selection and Selective Sweeps}

@@

%A predicted consequence of a reduction in Ne is an accelerated rate of ¢xation of slightly deleterious substitu- tions, and a reduction in the rate of ¢xation of slightly advantageous mutations (Kimura 1983). Thus we might expect an increased rate of amino-acid replacement rela- tive to silent substitutions on evolving Y chromosomes (provided that su¤cient time has elapsed since the origin of the sex-chromosome system), and more ¢xations of non-preferred versus preferred synonymous substitutions at silent sites in exons (Charlesworth 1996a).

@@

% A further interesting aspect of the results from the S. latifolia Y chromosome is that they do not readily ¢t a recent selective sweep. Such an event would eliminate all diversity, so that existing SLY1 polymorphisms must have accumulated since the selective sweep and should thus mostly be at low frequencies (see table 1). However, all six SLY1 polymorphic sites found are non-singletons, and no signi¢cant deviations from neutrality are detectable by standard test statistics (Fu & Li 1993). However, this does not take into account the probable population subdivision in this species. Although a selective sweep eliminating variability throughout the worldwide population of this species seems unlikely, e¡ects of selective sweeps within local populations cannot be excluded.

%because the lack of recombination means that all genes on the Yof any given species share a common genealogy, which limits the amount of information that can be obtained from data on within-species variability.

%This raises the further issue of whether Y-chromosome degeneration is driven mainly by the accumulation of deleterious mutations on the Y or is due to processes turning down the expression of genes on an evolving Y.
%We have focused on the ¢rst of these possibilities, but it might well not completely explain Y degeneration. If mutations of relatively minor ¢tness e¡ects, with Nes not much bigger than 1, become ¢xed more rapidly on a degenerating Ychromosome than on the X, selection pres- sure would develop to reduce the transcription of Y genes, provided that this was compensated for in the hetero- gametic sex by an increased activity of their counterparts on the X.

@@

The observation of widespread degeneration and diversity loss on Y chromosomes illustrates the importance of recombination for maintaining fitness and the genetic variability needed for adaptation (Maynard Smith 1978; Kondrashov 19 93; Barton & Charlesworth 19 9 8). The accumulation of deleterious mutations and extensive loss of diversity on a young plant Y chromosome revealed by our study provides a clear example of the reduced ability of selection to eliminate deleterious alleles from large genomic regions that lack recombination.
